Some examples in extended logic counting classes

### 2015-10-09 tex/examples.tex
9 oct

\subsection{{\sc \#Disj-Horn-Sat}}
We define the decision problem
\[
	\textsc{Disj-Horn-Sat} = \{\Phi \mid \Phi \mbox{ is a disjunction of Horn formulas and $\Phi$ is satisfiable}\},
\]
and the counting problem {\sc \#Disj-Horn-Sat} as a function that counts all satisfying assignments to a formula $\Phi$ which is a disjunction of Horn formulas. We will now show that {\sc \#Disj-Horn-Sat} $\in\EtwoE$.

>>>>

\subsection{{\sc \#Disj-Horn-Sat}}
We will now show that {\sc \#Disj-Horn-Sat} $\in\EtwoE$.

### 2015-10-09 tex/first_h.tex
9 oct

\section{Some properties of $\E1$}

We define the vocabulary $\L = \{ S_1,\dots,S_t, \leq \}$, where $S_1,\dots,S_t$ have arity $b_1,\dots,b_t$. Let
\begin{eqnarray*}
	\Truc &=& \{\A \mid \A \text{ is an } \L \text{-structure with a finite domain } A \text{ such that} \\
	&& \leq \text{ is interpreted as a total order for } A \}.
\end{eqnarray*}
We also define a set of second order variables ${\cal X} = \{ X_i \mid i\in\N \}$ where $X_i$ has arity $a_i$, and for every $n \in \N$ there are infinite variables in ${\cal X}$ of arity $n$. A quantifier-free $\L$-formula is defined by the following grammar:
\begin{eqnarray*}
	\varphi &::=& x = y \ \mid \ S_i(x_1,\dots,x_{b_i}), i \in \{1,\dots,t\} \ \mid \ x \leq y \ \mid \\
	&& X_i(x_1,\dots,x_{a_i}), i\in\N \ \mid \\ 
	&& (\neg \varphi) \ \mid \ (\varphi \wedge \varphi) \ \mid \ (\varphi \vee \varphi),
\end{eqnarray*}
where $x,y$ and $x_i$ are first order variables for every $i$. We now define an {\sc FO}-extended quantifier-free $\L$-formula as follows:
\begin{eqnarray*}
	\varphi &::=& \alpha, \alpha \text{ is an FO-formula over } \L  \ \mid \\
	&& X_i(x_1,\dots,x_{a_i}), i\in\N \ \mid \ \\
	&& (\neg \varphi) \ \mid \ (\varphi \wedge \varphi) \ \mid \ (\varphi \vee \varphi).
\end{eqnarray*}
Let $\Y = (Y_1,\dots,Y_q)$ be a tuple of second-order variables of arity $c_1,\ldots,c_q$, and let $\y$ be a tuple of first order variables. For every $\L$-formula $\psi(\Y,\y)$, we define the function $f_{\psi(\Y,\y)}:\Truc \to \N$ as follows:
\begin{eqnarray*}
	f_{\psi(\Y,\y)}(\A) &=& \mid \{ \langle \P, \e \rangle \mid \A \models \psi(\P,\e) \} \mid,
\end{eqnarray*}
for every $\A = \langle A, \S^{\A}, \leq^{\A} \rangle \in \Truc$, where $\P = (P_1,\ldots,P_q)$ is a tuple of predicates of arity $c_1,\ldots,c_q$, $P_i \subseteq A^{c_i}$ for every $i \in \{1,\ldots,q\}$, and $\e$ is a tuple of elements from $\A$.\\

A function $f:\Truc \to \N$ is in $\Eo$ if there exists an {\sc FO}-extended quantifier-free $\L$-formula $\varphi(\Y,\y)$ such that $f = f_{\varphi(\Y,\y)}$.

Similarly, a function $f:\Truc \to \N$ is in $\E1$ if there exists an {\sc FO}-extended quantifier-free $\L$-formula $\varphi(\x,\Y,\y)$ such that $f = f_{\exists \x \: \varphi(\x,\Y,\y)}.$\\

@@

		\item There is only second-order free variables in $\psi$. Let $\psi = \exists \x\,\varphi(\x,\X)$, where $\x = (x_1,\ldots,x_c)$ and $\X = (X_1,\ldots,X_r)$. Given a structure $\A = \langle A, \S^{\A}, \leq^{\A} \rangle\in\Truc$, we generate a DNF-formula $\Phi = h(\A)$ as follows:

>>>>

\section{Extended Logic Counting Classes}

We define the vocabulary $\L = \{ S_1,\dots,S_t, \leq \}$, where $S_1,\dots,S_t$ have arity $b_1,\dots,b_t$. Let
\begin{eqnarray*}
	\Truc &=& \{\A \mid \A \text{ is an } \L \text{-structure with a finite domain } A \text{ such that} \\
	&& \leq \text{ is interpreted as a total order for } A \}.
\end{eqnarray*}
We also define a set of second order variables ${\cal X} = \{ X_i \mid i\in\N \}$ where $X_i$ has arity $a_i$, and for every $n \in \N$ there are infinite variables in ${\cal X}$ of arity $n$. A quantifier-free $\L$-formula is defined by the following grammar:
\begin{eqnarray*}
	\varphi &::=& x = y \ \mid \ S_i(x_1,\dots,x_{b_i}), i \in \{1,\dots,t\} \ \mid \ x \leq y \ \mid \\
	&& X_i(x_1,\dots,x_{a_i}), i\in\N \ \mid \\ 
	&& (\neg \varphi) \ \mid \ (\varphi \wedge \varphi) \ \mid \ (\varphi \vee \varphi),
\end{eqnarray*}
where $x,y$ and $x_i$ are first order variables for every $i$. For every logic $\mathcal{F}$, we define an $\mathcal{F}$-extended quantifier-free $\L$-formula as follows:
\begin{eqnarray*}
	\varphi &::=& \alpha, \alpha \text{ is an $\mathcal{F}$-formula over } \L  \ \mid \\
	&& X_i(x_1,\dots,x_{a_i}), i\in\N \ \mid \ \\
	&& (\neg \varphi) \ \mid \ (\varphi \wedge \varphi) \ \mid \ (\varphi \vee \varphi).
\end{eqnarray*}
Let $\Y = (Y_1,\dots,Y_q)$ be a tuple of second-order variables of arity $c_1,\ldots,c_q$, and let $\y$ be a tuple of first order variables. For every $\L$-formula $\psi(\Y,\y)$, we define the function $f_{\psi(\Y,\y)}:\Truc \to \N$ as follows:
\begin{eqnarray*}
	f_{\psi(\Y,\y)}(\A) &=& \mid \{ \langle \P, \e \rangle \mid \A \models \psi(\P,\e) \} \mid,
\end{eqnarray*}
for every $\A = \langle A, \S^{\A}, \leq^{\A} \rangle \in \Truc$, where $\P = (P_1,\ldots,P_q)$ is a tuple of predicates of arity $c_1,\ldots,c_q$, $P_i \subseteq A^{c_i}$ for every $i \in \{1,\ldots,q\}$, and $\e$ is a tuple of elements from $\A$.\\

\subsection{The First {\sc FO}-Extended Counting Hierarchy}

We define syntactically the classes $\Eie$ and $\Eiu$ according to the following grammar:
\begin{enumerate}
	\item $\Eezero$:
	\[
		E_0 ::= \varphi \mid \varphi \mbox{ is an {\sc FO}-extended quantifier-free $\L$-formula }.
	\]
	\item $\Euzero$:
		\[U_0 ::= E_0.\]
	\item $\Eipe$:
	\[	E_{i+1} ::= U_i \ \mid \ \exists x \, E_{i+1}.\]
	\item $\Eipu$:
		\[U_{i+1} ::= E_i \ \mid \ \forall x \, U_{i+1}.\]
\end{enumerate}

A function $f$ is in $\Eie$ (resp. $\Eiu$) if there is an $\L$-formula $\varphi$ defined by the grammar $E_i$ (resp. $U_i$) such that $f = f_{\varphi}$.

@@

		\item There are only second-order free variables in $\psi$. Let $\psi = \exists \x\,\varphi(\x,\X)$, where $\x = (x_1,\ldots,x_c)$ and $\X = (X_1,\ldots,X_r)$. Given a structure $\A = \langle A, \S^{\A}, \leq^{\A} \rangle\in\Truc$, we generate a DNF-formula $\Phi = h(\A)$ as follows:

### 2015-10-09 tex/horn_h.tex
9 oct

>>>>

We define the decision problem
\[
\textsc{Disj-Horn-Sat} = \{\Phi \mid \Phi \mbox{ is a disjunction of Horn formulas and $\Phi$ is satisfiable}\},
\]
and the counting problem {\sc \#Disj-Horn-Sat} as a function that counts all satisfying assignments to a formula $\Phi$ which is a disjunction of Horn formulas.

\begin{theo}
	$\textsc{Disj-Horn-Sat}$ is hard for $\EtwoE$ under parsimonious reductions. 
\end{theo} 
\begin{proof}
	Let $f$ be an arbitrary function in $\EtwoE$ and let $f_{\textsc{Disj-Horn-Sat}}$ be the function that characterizes {\sc \#Disj-Horn-Sat}. We will now show a function $h:\Truc\to L(P)$ such that for every $\A\in\Truc$, it holds that $f(\A) = f_{\textsc{Disj-Horn-Sat}}(h(\A))$, and which can be computed in polynomial time. Let $\psi$ be such that $f = f_{\psi}$. We separate the proof in three cases:\begin{enumerate}
		\item All free variables in $\psi$ are of first order. Let $\psi = \exists \x \forall \y\,\varphi(\x,\y,\z)$ where $\z = (z_1,\ldots,z_d)$. Given a structure $\A = \langle A, \S^{\A}, \leq^{\A} \rangle\in\Truc$, we generate a propositional formula $\Phi = h(\A)$. First we notice that this case is trivial since we can calculate this function in polynomial time, by checking for each $\e\in A^d$ if $\A\models\exists\x\forall \y\,\varphi(\x,\y,\e)$. Let $n$ be the number of such assignments to $\z$. Let $\p$ be an $n$-tuple of propositional variables where $\p = (p_1,\ldots,p_n)$. The Disj-Horn-Sat formula we return is
		\[
		\Phi = \bigvee_{i = 1}^n \neg p_1 \wedge \cdots \wedge \neg p_{i-1} \wedge p_i \wedge \neg p_{i+1} \wedge \cdots \wedge \neg p_n
		\]
		which has exactly $n$ possible assignments. Note that each $p_i$ and $\neg p_i$ is a Horn clause itself.

		\item There are only second-order free variables in $\psi$. Let $\psi = \exists \x\forall \y\,\varphi(\x,\y,\X)$, where $\x = (x_1,\ldots,x_c)$, $\y = (y_1,\ldots,y_b)$, $\X = (X_1,\ldots,X_r)$ and each $X_{\ell}$ has arity $c_{\ell}$. Given a structure $\A = \langle A, \S^{\A}, \leq^{\A} \rangle\in\Truc$, we generate a Disj-Horn-Sat-formula $\Phi = h(\A)$ as follows:

		First, Let $\varphi(\x,\y,\X) = \bigwedge_{i = 1}^k \varphi_{i}(\x,\y,\X)$ where, without loss of generality, each $\varphi_{i}(\x,\y,\X)$ is of the form
		\[
		\varphi_{i}(\x,\y,\X) = X_s(\u_{s}) \vee \neg X_{t_1}(\u_{t_1}) \vee \cdots \vee \neg X_{t_n}(\u_{t_n}) \vee \alpha_i(\x,\y),
		\]
		where $\u_{s}$ and every $\u_{t_j}$ have the corresponding number of variables in $(\x,\y)$, and $\alpha(\x,\y)$ is a FO-formula. For the rest of the proof, $X_s(\u_s)$ can be absent and it does not affect the proof. Note that this formula is equivalent to
		\[
		\neg\alpha_i(\x,\y) \to (X_s(\u_{s}) \vee \neg X_{t_1}(\u_{t_1}) \vee \cdots \vee \neg X_{t_n}(\u_{t_n})),
		\]

		Then, for each $\a\in A^c$ let $\Gamma_i^{\a} = \{\b\in A^{b}\mid \A\models\neg\alpha(\a,\b)\}$. Note that this set can be generated in polynomial time. For each $\ell\in\{1,\ldots,r\}$, we generate a propositional variable $p^{\ell}_{\d}$ for each $\d\in A^{c_{\ell}}$. Let $\p$ be the tuple of all such variables. Also, let $\rho_{\ell}(\x,\y) = \u_{\ell}$, and $\rho_{\ell}(\d)$ be the tuple that results of reordering $\d$ using the reordering of the variables of $(\x,\y)$ in $\u_{\ell}$.

		For each $\a\in A^c$ we generate a Horn formula
		\[
		\Phi_{\a}^i = \bigwedge_{\b\in\Gamma_i} p^{s}_{\rho_s(\a,\b)} \vee \neg p^{t_1}_{\rho_{t_1}(\a,\b)} \vee \cdots \vee \neg p^{t_n}_{\rho_{t_n}(\a,\b)}.
		\]
		and finally we return $\Phi = \bigvee_{\a\in A^c} \bigwedge_{i = 1}^k \Phi_{\a}^i$.

		Note that for each assignment $\P$ to $\X$ such that $\A\models\exists\x\forall\y\,\varphi(\x,\y,\P)$ there is a corresponding assignment $\sigma$ to $\p$ where $\sigma(p^{\ell}_{\a}) = 1$ if and only if $\a\in P_{\ell}$. We can conclude that $f(\A) = f_{\textsc{Disj-Horn-Sat}}(\Phi)$.

		\item 
	\end{enumerate}
\end{proof}
